---
title: "BN - Assignment 2"
author: "Thomas Rost"
date: "Saturday, December 13, 2014"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
bibliography: C:\Users\Trost\Copy\Nijmegen\Storage\Docs\Thomas\Mendeley_Bib\library.bib
---

\newpage


\section{test}



First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell

# assignment


[-@Steele1999]
@Steele1999

## comparison of two learning algos

### data discretization
data discretization: number of arcs, changes in arc direction,...
three discretizations in the iris data with different number of bins (B.3). Do this for both classes (constraing based oder search and score). Comppare the results

### size of dataset
consider subsets of BC-dataset with different number of samples. both classes of learning algios, compare results

## comparison to manually constructed bayes network

NHL dataset, liad column classes as factor (A.4)

missing data : probably median since mean desn't make sense for factors.

check for differences, unterschiedlcihe zahl nodes is ok.

### define measures for quality of a learning algorithm in terms of a known bayes network structure, motivate definition

###use measures to evaluate both classes of learning algs for NHL dataset

### compare margial prob distributions of learnt NHL bN with those of the manually coinstructed ones

## diagnostic BN structures

leanr bayes network from BN using sas or cb algo

## develop special purpose BN (NBN or TAN, see .2)

## compare BN and SPBN with manually constructed network in terms of network structure, goodnes of fit scores (lkikelohood,...) and accuracy measures such as misclassification error and ROC (B.2)

library(ROCR),
\footnote{Tobias Sing, Oliver Sander, Niko Beerenwinkel, Thomas Lengauer.
ROCR: visualizing classifier performance in R.
Bioinformatics 21(20):3940-3941 (2005).}
\cite{Sing2005}
@Sing2005
 








# Ideas:


```{r}
library(Rgraphviz)
library(gRain)
library(bnlearn)
library(caTools)
data(asia)

net = hc(asia)
plot(net)
net <- set.arc(net,"A","T")

plot(net)
fitted = bn.fit(net,asia)
fitted

score(net,asia)

netgs = gs(asia)
netgs_directed = cextend(netgs)
netgs_directed
plot(netgs)
plot(netgs_directed)


netnb = naive.bayes(asia,"T",names(asia)[c(1:2,4:8)])
plot(netnb)
```

Do Cross validation!

```{r}
netcv = bn.cv(asia,net,loss ="pred", k = 3, loss.args = list(target="T"),debug = T)
```




#references
\twocolumn